---
title: "A1 Task 2: PCA Exploration"
author: "Nicole Schmidt"
date: "1/21/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# Add packages and read in data 
library(tidyverse)
library(janitor)
library(here)

#for PCA
library(ggfortify)

# For ggplot customization
library(readxl)
library(gghighlight)
library(lubridate)
library(plotly)

world_env_var <- read_csv("world_env_vars.csv")
 

```

#### 1. Introduction

The data used for this project is miscellaneous environmental and climatic variables that are are described within countries. For the purpose of this project I am exploring the variance between elevation, cropland cover, tree canopy cover, mean annual rainfall, and mean annual temperature  This data was compiled by Zander Venter and acquired from Google Earth Engine.


```{r, message = FALSE, warning = FALSE}

#Wrangle data to keep variables of interest (elevation, crop cover, tree canopy cover, annual mean temperature, and annual mean rainfall). Drop na values and scale data 

# is it better to keep all countries or is it possible to filter and explore?
#within the top 5 carbon emitting countries, China, United States of America, India, Russia, and Japan (Union of Concerned Scientists, 2020). citation: https://www.ucsusa.org/resources/each-countrys-share-co2-emissions

#PCA:
world_climate_pca <- world_env_var %>% 
  select(cropland_cover, tree_canopy_cover, elevation, ends_with("annual")) %>% 
  drop_na() %>%
  scale() %>%  #values have been scaled
  prcomp() #now a list containing information 

world_climate_pca$rotation #brings up loading for variables 

world_climate_complete <- world_env_var %>% 
  drop_na(cropland_cover, tree_canopy_cover, elevation, ends_with("annual")) 

#Make a Biplot: 
autoplot(world_climate_pca,
         data = world_climate_complete,
         colour = 'Country',
         loadings = TRUE,
         loadings.label = TRUE,
         loadings.colour = "black",
         loadings.label.colour = "black",
         loadings.label.vjust = 1,
         loadings.label.hjust = -.15)+
  theme_bw()+
  theme(legend.position = 'none')
# autopilot function understands outcomes from data analysis to create plot that it assumes you want to make 



```

#### A brief summary (nicely formatted and professionally written bulletpoints are fine) highlighting some major takeaways from your PCA that can be gleaned from the biplot
